Hippocampal segmentation with MAGeT

### 2013-01-16 paper.Rnw
Updated the intro organisation

\documentclass[twocolumn]{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable }
\usepackage[square,numbers]{natbib}
\setlength{\marginparwidth}{1.2in}
\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
\newcommand{\todo}[1]{\comment{TODO #1}}
\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}

@@

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo{Finish this... }
\end{abstract}

@@

\subsection{MAGeT Brain Algorithm}
\todo{Make sure we explain what MAGeT stands for}
\todo{reference Chakravarty2012}
In this paper, we use the term {\it atlas} to mean any manually segmented MR
image, and the term {\it atlas library} to mean a set of such images.  We use
the term {\it template} to refer to any MR image, and associated labelling,
used to segment another image, and the term {\it template library} to refer to
a set of such images.  An atlas library may be used as a template library but,
as we will discuss, a template library may also be composed of images with
computer generated labellings.

The segmentation approach we propose is best understood as an extension of
basic multi-atlas segmentation \citep{Collins2010}.  In multi-atlas segmentation,
an atlas library and unlabelled MR images are given as input.  Every atlas
image is nonlinearly registered to each unlabelled image, and then each atlas'
labels are propagated via the resulting transformations.  These labels are then
fused to produce a single, definitive segmentation by some label fusion method
(e.g.  voxel-wise majority vote).

Our extension adds a preliminary stage in which a template library is
constructed from input images, and used in place of an atlas library in the
standard multi-atlas-based method.  To create the template library, labels from
each atlas image are propagated to each template library image via the
transformation resulting from a non-linear registration between pair of images.
As a result, each template library image has a label from each atlas.
Basic multi-atlas segmentation is then used to produce segmentations for the
entire set of unlabelled images (including those images used in the template
library).

@@

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the MAGeT Brain algorithm}
  \label{pseudocodesdf}
  \begin{algorithmic}
    \Function{MultiAtlas}{Templates, Subjects}
      \ForAll{$subject$}
        \ForAll{$template$}
          \State propagate all labels for template to subject space
          \State store subject labels
        \EndFor
        \State fuse subject labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a subject to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\begin{figure*}
<<ADNI-1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 + MMSE, 
               data = adnimerge, subset = VISCODE == 'bl',   
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=TRUE,
      caption="ADNI-1 1.5T Screening demographics")
@
\end{figure*}

@@

\subsubsection{SZ First Episode Patients}
\todo{include image characteristics}
\begin{figure*}
<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", size="scriptsize", 
      caption="Schizophrenia First Episode Patient Demographics")
@
\end{figure*}

@@

We used the following command line when running the ANTS command, 
\begin{verbatim}
  mincANTS 3 -m PR[target_file.mnc,source_file.mnc,1,4] 
   --number-of-affine-iterations 10000x10000x10000x10000x10000 
   --affine-gradient-descent-option 0.5x0.95x1.e-4x1.e-4
   --use-Histogram-Matching --MI-option 32x16000
   -r Gauss[3,0] -t SyN[0.5] -i 100x100x100x20
   -o transformation.xfm
 \end{verbatim}
These settings were adapted from the "reasonable starting point" given in the
ANTS manual. \todo{include a reference to manual?}

@@

\begin{table*}
<<ADNI-69-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c("295","413","619","685","729","782","938","954",
                 "1018","1155","907","981","222","324","448","546",
                 "553","572","602","610","814","1341","675","681",
                 "731","1130","41","68","70","101","249","293","316",
                 "414","698","1206","1222","1339","751","1030","3",
                 "5","10","16","22","53","168","183","241","1205",
                 "328","1095","991","213","159","337","343","642",
                 "753","1109","1217","29","56","1188","1293","149",
                 "382","431","1202")
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 + MMSE, 
               data = adnimerge, subset = VISCODE == 'bl' & RID %in% adni_69_RIDs, 
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=TRUE,
      caption="ADNI-1 cross-validation subset demographics")
@
\end{table*}

>>>>

\documentclass[twocolumn]{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable }
\usepackage[square,numbers]{natbib}

%draft mode
\usepackage[colorlinks]{hyperref}
\usepackage[colorinlistoftodos, textwidth=4cm, shadow]{todonotes}
\usepackage[displaymath, tightpage]{preview}

\setlength{\marginparwidth}{1.2in}

%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}

@@

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo[inline]{Finish this... }
\end{abstract}

@@

\subsection{MAGeT Brain Algorithm}
\todo{Make sure we explain what MAGeT stands for}
\todo{reference Chakravarty2012}

In this paper, we use the term {\it atlas} to mean any manually segmented MR
image, and the term {\it atlas library} to mean a set of such images.  We use
the term {\it template} to refer to any MR image, and associated labelling,
used to segment another image, and the term {\it template library} to refer to
a set of such images.  An atlas library may be used as a template library but,
as we will discuss, a template library may also be composed of images with
computer generated labellings.

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion\citep{Collins2010}.  We will refer to
this as {\em basic multi-atlas segmentation}.  The primary steps are as
follows.  An atlas library and unlabelled MR images are given as input.  Each
atlas image is nonlinearly registered to each unlabelled image, and then each
atlas' labels are propagated via the resulting transformations.  The resulting
labels are fused to produce a single, definitive segmentation by some label
fusion method (for exampl, by voxel-wise majority voting).

The segmentation approach we propose is best understood as an extension of
MAGeT brain adds a preliminary stage to this process in which the template
library is constructed rather than given as input.  As before, MAGeT brain
accepts an atlas library and unlabelled MR images as input.  Images for the
template library are selected from

To create the template library, labels from each atlas image are propagated to each template
library image via the transformation resulting from a non-linear registration
between pair of images.  As a result, each template library image has a label
from each atlas.  Basic multi-atlas segmentation is then used to produce
segmentations for the entire set of unlabelled images (including those images
used in the template library).

@@

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the MAGeT Brain algorithm}
  \label{pseudocodesdf}
  \begin{algorithmic}
    \Function{BasicMultiAtlasSegmentation}{Templates, Subjects}
      \ForAll{$subject$}
        \ForAll{$template$}
          \State propagate all labels for template to subject space
          \State store subject labels
        \EndFor
        \State fuse subject labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a subject to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

<<ADNI-1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 + MMSE, 
               data = adnimerge, subset = VISCODE == 'bl',   
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",  caption="ADNI-1 1.5T Screening demographics")
@

@@

\subsubsection{SZ First Episode Patients}
\todo{include image characteristics}

<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", size="scriptsize", ctable=FALSE,
      caption="Schizophrenia First Episode Patient Demographics")
@

@@

We used the following command line when running the ANTS command, 
{\small
\begin{verbatim}
  mincANTS 3 -m PR[target_file.mnc,source_file.mnc,1,4] 
   --number-of-affine-iterations 10000x10000x10000x10000x10000 
   --affine-gradient-descent-option 0.5x0.95x1.e-4x1.e-4
   --use-Histogram-Matching --MI-option 32x16000
   -r Gauss[3,0] -t SyN[0.5] -i 100x100x100x20
   -o transformation.xfm
 \end{verbatim}
}
These settings were adapted from the "reasonable starting point" given in the
ANTS manual. \todo{include a reference to manual?}

@@

<<ADNI-69-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c("295","413","619","685","729","782","938","954",
                 "1018","1155","907","981","222","324","448","546",
                 "553","572","602","610","814","1341","675","681",
                 "731","1130","41","68","70","101","249","293","316",
                 "414","698","1206","1222","1339","751","1030","3",
                 "5","10","16","22","53","168","183","241","1205",
                 "328","1095","991","213","159","337","343","642",
                 "753","1109","1217","29","56","1188","1293","149",
                 "382","431","1202")
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 + MMSE, 
               data = adnimerge, subset = VISCODE == 'bl' & RID %in% adni_69_RIDs, 
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption="ADNI-1 cross-validation subset demographics")
@
